---
title: "Cyclistic Bike-Share Case Study"
author: "Philemon John"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cosmo
    highlight: tango
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This project is the **Google Data Analytics Capstone Case Study**: *How does a bike-share navigate speedy success?*  
I applied the six-step data analysis process: **Ask, Prepare, Process, Analyze, Share, and Act**.  

---

## 1. Ask Phase

### Business Task
To understand **how annual members and casual riders use Cyclistic bikes differently**, and design strategies to **convert casual riders into annual members**.

### Key Stakeholders
- **Cyclistic marketing team** – plans campaigns based on rider behavior.  
- **Cyclistic executive team** – approves membership strategy.  
- **Data analytics team** – supports data-driven decision-making.

---

## 2. Prepare Phase

### Data Source
The dataset comes from **Cyclistic’s trip data** made available by Motivate International Inc.  
It contains details of rides from the **past 12 months** and includes:

- Ride ID  
- Bike type  
- Start and end timestamps  
- Station details (name, ID, latitude, longitude)  
- User type (member or casual)

Data is publicly available and anonymized, ensuring privacy compliance.

```{r load-packages, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
knitr::opts_chunk$set(echo = TRUE)
```

# Load all CSV files in 'data' folder
```{r}
all_trips <- list.files(path = "C:/data/Cyclistic_Project/data", pattern = "*.csv", full.names = TRUE) %>%
lapply(read_csv) %>%
bind_rows()

glimpse(all_trips)
```

## 3.Process Phase

Data Cleaning

Steps performed:

1.Removed duplicates and missing values where necessary.

2.Converted date-time columns to proper format.

3.Created new columns for ride length, day of week, month, and hour.

```{r}
# Read and clean data first
all_trips <- all_trips %>%
   # Step 1: make column names consistent
  mutate(
    started_at = as_datetime(started_at),
    ended_at = as_datetime(ended_at),
    ride_length = as.numeric(difftime(ended_at, started_at, units = "mins")),
    day_of_week = wday(started_at, label = TRUE),
    month = month(started_at, label = TRUE, abbr = FALSE),
    hour = hour(started_at)
  ) %>%
  filter(ride_length > 0) %>%
  clean_names()   
```


4. Analyze Phase

We explored how ride duration, day of the week, and number of rides vary by user type.

4.1 Average Ride Duration by User Type

```{r}
avg_ride_length <- all_trips %>%
group_by(member_casual) %>%
summarise(mean_ride = mean(ride_length), median_ride = median(ride_length), max_ride = max(ride_length))

avg_ride_length
```

4.2 Average Ride Duration by Day of Week

```{r}
avg_ride_day <- all_trips %>%
group_by(member_casual, day_of_week) %>%
summarise(mean_ride = mean(ride_length), .groups = 'drop')

avg_ride_day
```

4.3 Number of Rides by Day of Week

```{r}
rides_per_day <- all_trips %>%
group_by(member_casual, day_of_week) %>%
summarise(number_of_rides = n(), .groups = 'drop')

rides_per_day
```

5. Share Phase

Visualization: Average Ride Duration

```{r}
ggplot(avg_ride_length, aes(x = member_casual, y = mean_ride, fill = member_casual)) +
geom_col(show.legend = FALSE) +
labs(title = "Average Ride Duration by User Type",
x = "User Type",
y = "Average Ride Duration (minutes)") +
theme_minimal()
```

Visualization: Number of Rides by Day of Week

```{r}
ggplot(rides_per_day, aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
geom_col(position = "dodge") +
labs(title = "Number of Rides by Day of Week and User Type",
x = "Day of Week",
y = "Number of Rides") +
theme_minimal()
```

Visualization: Average Ride Duration by Day

```{r}
ggplot(avg_ride_day, aes(x = day_of_week, y = mean_ride, color = member_casual, group = member_casual)) +
geom_line(size = 1.2) +
geom_point(size = 2) +
labs(title = "Average Ride Duration by Day of Week",
x = "Day of Week",
y = "Average Ride Duration (minutes)") +
theme_minimal()
```

6. Act Phase
Key Findings

Casual riders have longer average ride times (≈ 23 mins) compared to members (≈ 12 mins).

Members ride more frequently, particularly on weekdays, indicating commute patterns.

Casual riders peak on weekends, suggesting leisure use.

Ride durations for casual users spike during weekends and midday hours.

Recommendations

Convert casuals to members via:

Weekend membership promotions.

Seasonal or tourist membership options.

Reward programs for repeated rides.

Corporate partnerships: Offer discounted memberships for employees.

In-app marketing: Show casual users their potential savings if they became members.

7. Conclusion

This analysis reveals clear behavioral patterns distinguishing casual riders from members.
By leveraging these insights, Cyclistic can craft targeted marketing campaigns to boost membership and revenue.

8. Next Steps

Analyze station-level data to understand popular start and end locations.

Incorporate weather data to see its influence on riding behavior.

Build predictive models to forecast peak riding times.

Prepared by Philemon John
Data Analyst
