---
title: "Cyclistic Bike-Share Analysis"
author: "Philemon John"
output: html_document
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
Cyclistic aims to convert casual riders into annual members. This report summarizes key insights from usage data and recommends marketing actions based on ride behavior patterns.


## Data Summary
The dataset contains over 5.5 million rides from October 2023 to October 2024, with information on user type, duration, day of week, and rideable type.

## Key Findings

### 1. Average Ride Duration
```{r}
# Insert your chart from previous step
all_trips %>%
  group_by(member_casual) %>%
  summarise(mean_ride = mean(ride_length)) %>%
  ggplot(aes(x = member_casual, y = mean_ride, fill = member_casual)) +
  geom_col() +
  labs(title = "Average Ride Duration by User Type",
       x = "User Type", y = "Average Duration (minutes)") +
  theme_minimal()
```

Insight:
Casual riders average 23 minutes per ride, nearly twice as long as members (12 minutes), showing that casual riders often ride for recreation or leisure.

### 2. Ride Patterns by Day of Week

```{r}
all_trips %>%
group_by(member_casual, day_of_week) %>%
summarise(number_of_rides = n(), .groups = 'drop') %>%
ggplot(aes(x = day_of_week, y = number_of_rides, fill = member_casual)) +
geom_col(position = "dodge") +
labs(title = "Number of Rides by Day and User Type",
x = "Day of Week", y = "Number of Rides") +
theme_minimal()
```

Insight:
Casual riders mostly use bikes on weekends, while members ride consistently throughout the week — suggesting weekday commuting patterns.

### 3.Average Ride Duration by Day of Week

```{r}
avg_ride_day <- all_trips %>%
  group_by(member_casual, day_of_week) %>%
  summarise(mean_ride = mean(ride_length, na.rm = TRUE), .groups = 'drop')

ggplot(avg_ride_day, aes(x = day_of_week, y = mean_ride, color = member_casual, group = member_casual)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Average Ride Duration by Day of Week",
       x = "Day of Week",
       y = "Average Ride Duration (minutes)") +
  theme_minimal()
```
Insight:
Casual riders’ ride durations spike on weekends, while members maintain consistent short rides across the week.

### 4.Rides by Hour of Day
```{r}
all_trips$hour <- lubridate::hour(all_trips$started_at)

rides_by_hour <- all_trips %>%
  group_by(member_casual, hour) %>%
  summarise(rides = n(), .groups = 'drop')

ggplot(rides_by_hour, aes(x = hour, y = rides, color = member_casual)) +
  geom_line(size = 1.2) +
  labs(title = "Rides by Hour of Day",
       x = "Hour of Day (24h)",
       y = "Number of Rides") +
  theme_minimal()
```

Insight:
Members peak around 7–9 AM and 5–6 PM (commute hours).
Casual riders peak midday and late afternoon, typical of leisure activities.

# Insights Summary:
# 1. Members take shorter, consistent rides mainly on weekdays (commuting).
# 2. Casual riders take longer rides, mainly on weekends (leisure).
# 3. Casual riders’ peak hours are mid-day, while members’ peaks are during commute hours.
# 4. Members generate steady usage throughout the week, showing loyalty.
# 5. Casual riders can be targeted with weekend offers, loyalty points, or membership discounts.


##Recommendations (Act Phase)

### 1. Weekend Membership Promotions:
Offer discounted weekend passes to encourage casual riders to transition to annual plans.

### 2. Commuter Conversion Campaign:
Target casual riders who ride frequently on weekdays — promote benefits of unlimited rides and cost savings.

### 3. Mobile App Nudges:
Use in-app notifications to show casual riders how much they’d save if they subscribed.

### 4. Dockless Feature Emphasis:
Highlight flexibility for members using electric bikes that don’t require docking stations.


## Summary Tables
```{r}
ggsave("avg_ride_duration.png", width = 6, height = 4)
ggsave("rides_by_day.png", width = 6, height = 4)
write.csv(avg_ride_day, "average_ride_day_summary.csv", row.names = FALSE)
```

##Conclusion

Data shows distinct behavioral differences between casual and member riders. By aligning marketing efforts with casual riders’ weekend habits and leisure interests, Cyclistic can increase conversions to annual membership.






